---
title: 'Tidy Tuesday Shiny App: Nuclear Explosions'
author: Juan Manuel Canovas
date: '2019-12-01'
slug: tidy-tuesday-shiny-app-nuclear-explosions
categories: []
tags: []
output:
  blogdown::html_page:
    number_sections: true
    toc: true
---


# Introduction  

It’s been a long time since my last post, but I’m here again: this time with a Shiny App. Shiny is defined as a *"Web Application Framework for R".*   

From `?shiny`:

>> *Shiny makes it incredibly easy to build interactive web applications with R. Automatic "reactive" binding between inputs and outputs and extensive prebuilt widgets make it possible to build beautiful, responsive, and powerful applications with minimal effort.*

Although you will write them in R, Shiny Apps are a whole new world. One word in the definition above deserves special attention, specially if you don’t have experience in web development: **reactive**. 

**Reactivity: The key idea of reactive programming is to specify a graph of dependencies so that when an input changes, all related outputs are automatically updated. Also, [check this 2 minutes video.](https://vimeo.com/rstudioinc/review/131218530/212d8a5a7a/#t=47m25s)**  

It turns out reactivity is a fundamental concept in Shiny development. In fact, most powerful shiny apps relies heavily on this concept. Although mastering the concept will take some time, investing time and effort to grasp its foundations will pay off soon.

# The data  

## Brief description  

The data is from Stockholm International Peace Research Institute and shows the annual numbers of nuclear tests conducted
since 1945. Addtional information concerning the tests is also presented. While the list draws on a wide range of sources, the core data are those compiled by FOA (Swedish National Defence Research Institute).   

More information on this topic and related datasets can be found on this [article](https://ourworldindata.org/nuclear-weapons), by Max Roser and Mohamed Nagdy.

Before moving on, let's load some libraries we will need to work with the data.   

```{r, message=FALSE, warning=FALSE}
# required libraries
library(plyr)
library(dplyr)
library(tidyr)
library(stringr)
```

We now can import the data:  

```{r}
# import data
ne_original <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-08-20/nuclear_explosions.csv")
```


## Sources  

[GitHub repository](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-08-20) and original sources  


# What I learned along the way

## The use of `%in%` from base R   

This in an interface to the `match()` function, which returns a vector of the positions of matches of its first argument in its second. Let's see an example (taken from `?'%in%'`):

```{r, echo=TRUE}
# create two vectors
a <- 1:5
b <- 3:8

# print vector "a"
a

# print vector "b"
b

# check which elements from vector "a" are also in vector "b"
a %in% b
```


## The use of `mapvalues()` from {plyr}  

There are some regions that have been written differently, possibly due to the fact that the data is coming from different sources, as we can see below:  

```{r}
# "UKRAINE" in the USSR is written as "UKEAINE" and "UKRAINE"
ne_original %>% 
  select(country, region, source) %>% 
  filter(str_detect(region, "^UK"))

# "FANGATAUFAA" in France is written as "FANGATAUFAA" and "FANGATAUFA"
ne_original %>% 
  select(country, region, source) %>% 
  filter(str_detect(region, "^FANGATAUFA"))
```

An easy way to fix these cases is with function `mapvalues()` from {plyr} package. From `?mapvalues()` we can see that the function takes on the following arguments: `mapvalues(x, from, to, warn_missing = TRUE)`

* `x`: the factor or vector to modify  
* `from`: a vector of the items to replace  
* `to`: a vector of replacement values  
* `warn_missing`: print a message if any of the old values are not actually present in `x`

Let's see it in action:  

```{r, eval=FALSE}
# impute corrections for "UKRAINE"
ne <- ne %>% 
  mutate(region = mapvalues(region, 
                            from = "UKEAINE", 
                            to = "UKRAINE"))

# impute corrections for "FANGATAUFA"
ne <- ne %>% 
  mutate(region = mapvalues(region, 
                            from = c("FANGATAUFAA", "MUEUEOA"), 
                            to = c("FANGATAUFA", "MURUROA")))
```


## Timelines with {timevis} library 

based on javascript D3.js   


# Useful resources  

In no particular order, resources I found useful are:   

* [Mastering shiny](https://mastering-shiny.org/), a book by Hadley Wickham.
* [R Shiny app tutorial](https://www.youtube.com/watch?v=_0ORRJqctHE&list=PL6wLL_RojB5xNOhe2OTSd-DPkMLVY9DfB), a series of videos by Abhinav Agrawal.  
* [Shiny tutorial](https://shiny.rstudio.com/tutorial/), a series of videos by Garrett Grolemund.  
* [Shiny gallery](https://shiny.rstudio.com/gallery/), from RStudio.  
* [Building Shiny applications](https://cfss.uchicago.edu/notes/shiny/), by Dr. Benjamin Soltoff.  


# Conclusion  

You can check the Shiny app here:   
https://canovasjm.shinyapps.io/nuclear_explosions/

